### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/response/response.tex

\item{(3)} The literature study on monotone spline interpolation is incomplete and misses important work. As far as I am aware of, one of the most commonly used methods for shape-preserving interpolation is the quadratic spline method by Larry Schumaker (it does not only preserves monotonicity but also convexity!), at least in the spline approximation theory community:

\itemitem{} Larry L. Schumaker: On Shape Preserving Quadratic Spline Interpolation. SIAM Journal on Numerical Analysis, Vol. 20, No. 4 (1983), Pages 854-864.

}

The submission cites quadratic, cubic, and quartic previous works. The work by Larry Schumaker is now referenced in the attached update to the paper (section 1, paragraph 3). It is our understanding that one of the most widely used shape-preserving spline algorithms is a cubic spline methodology proposed by Fritsch and Carlson (known as PCHIP), which is included in many mathematical software libraries and cited in the submission.

@@

\item{(4)} The numerical section just shows few examples of data sets and the obtained spline results. The authors make some general statements about the algorithm like ``it generally provides aesthetically pleasing (low wiggle) initial estimates ... is preferred because it results in monotone quintic spline interpolants that are nearer to initial estimates'' but a deeper (more objective) discussion is lacking. Also a numerical comparison with other classical monotone spline methods is missing, at least the algorithm by Schumaker [1983] and also the one by Fritz/Carlson [1980]. This might help understand why the new method is of interest.

}

This is a reasonable criticism, but also easily addressed with minor revisions to the submission. While the reviewer may disagree with the language we used, there are objective reasons for the decisions made in this work in addition to the subjective comments made. Specifically, the quadratic facet model picks the minimum curvature local estimate for second derivatives which has the smallest magnitude second derivative, and the binary search identifies derivative values absolutely nearer to the initial estimates than the previous method of shrinking derivative values used in (1). These facts are now mentioned in the paper (section 4, paragraphs 2 and 3).

Comparisons with lower order methods are not appropriate for this work. There is no published mathematical software that also constructs a monotone quintic spline interpolant to data. If a user desires a monotonicity preserving $C^2$ approximation, this would be the only published software that achieves that goal. This paper is not intended to be a review and comparison of all available shape preserving spline packages. Including such comparisons would also require subjective decisions about what visuals to show, and those decisions might draw criticisms that distract from the unique purpose and achievement of this work (producing $C^2$ monotonicity preserving approximations to data). To emphasize that this work is not a replacement for previous works, but rather an additional method to consider, the first sentence of section 1, paragraph 5 that mentioned the importance of this work ``over'' previous methods was removed. The inherent difficulty of comparing this work with existing lower order methods is now mentioned in section 5, paragraph 5.

@@

\item{(5)} The most time-consuming part of the algorithm is the conversion to the B-spline representation. This has been done by solving a global linear system. It would be more efficient when using a local quasi-interpolation spline scheme with projection property.

\item{} Moreover, spline evaluation by explicitly computing the linear combination of B-splines is not efficient; instead use de Boor-Cox algorithm.

}

The reviewer may have missed the reason for the global linear system, from the submission (section 5, paragraph 4), ``It would be faster to construct splines over intervals independently ... the single linear system is chosen here for the decreased redundancy''. Comments in the paper and the Fortran source code indicated an option for returning the derivatives at the data points, so one can use either the piecewise polynomial or the B-spline representations. Further, the time for the banded system solve is $O(n)$, reasonable for $n$ data points. To be clear, the de Boor-Cox algorithm {\it is} used to evaluate the underlying B-splines (section 5, paragraph 1). The linear combination in the submission is of the (quintic) B-splines that define the interpolant.

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Incorporating some most recent comments from Dr. Watson and Thacker

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations with
large {\it wiggle}, or whose second derivative functions have large
$L2$ norm over the domain. In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

%% ===================================================================
%%                   Algorithm 1: QUADRATIC_FACET
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 1:} {
\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$

where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
local quadratic interpolant with minimum magnitude curvature.

@@

%% ===================================================================
%%                    Algorithm 2: IS_MONOTONE
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 2:}

{\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
$D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1) \bigr)$

where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.

@@

where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
points. Returns monotone quintic spline interpolant $Q(x)$ such that
$Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
intervals that $Y_i$ is monotone increasing (decreasing).

@@

where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
data points, and $Q(x)$ is a quintic spline interpolant such that
$Q(X_i) = Y_i$, $DQ(X_i) = u_i$, $D^2Q(X_i) = v_i$. Modifies
derivative values ($u$ and $v$) of $Q$ at data points to ensure
{\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
points, given a desired precision $\mu \in {\bbb R}$. proj$\bigl(w$,
int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
interval with endpoints $a$ and $b$.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large $L^2$ norm (approximations
with large {\it wiggle}). In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

%% ===================================================================
%%                   Algorithm 1: QUADRATIC_FACET
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 1:}
{\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$

\nobreak
where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
local quadratic interpolant with minimum magnitude curvature.

@@

%% ===================================================================
%%                    Algorithm 2: IS_MONOTONE
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 2:}
{\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
$D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1) \bigr)$

\nobreak
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.

@@

\nobreak
where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
points. Returns monotone quintic spline interpolant $Q(x)$ such that
$Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
intervals that $Y_i$ is monotone increasing (decreasing).

@@

\nobreak
where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
data points, and $Q(x)$ is a quintic spline interpolant such that
$Q(X_i) = Y_i$, $DQ(X_i) = u_i$, $D^2Q(X_i) = v_i$. Modifies
derivative values ($u$ and $v$) of $Q$ at data points to ensure
{\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
points, given a desired precision $\mu \in {\bbb R}$. proj$\bigl(w$,
int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
interval with endpoints $a$ and $b$.

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Incorporating some most recent comments from Dr. Watson and Thacker

The quadratic facet model produces a unique sensitivity to input
perturbation, as small changes in input may cause different quadratic
facets to be associated with a breakpoint, and thus different initial
derivative estimates can be produced. This phenomenon is depicted in
Figure 1. Despite this sensitivity, the quadratic facet model is still
preferred because it exactly captures local linear and quadratic
behavior while empirically producing final approximations with less
wiggle ($L2$ norm of the second derivative) than other methods. A
weighted harmonic mean estimate of first derivatives may be more
accurate when the underlying function changes at a rate greater than a
quadratic, but that method increases the sensitivity to small
perturbations in data and empirically results in quintic splines with
greater wiggle.

The binary search for a point on the monotone boundary in $(\tau_1$,
$\alpha$, $\beta$, $\gamma)$ space is performed because it results in
monotone quintic spline interpolants with derivative values that are
absolutely nearer to initial estimates than a search that strictly
shrinks derivative values. Given the initial derivative estimates have
desirable properties (capture low-order phenomena and are low wiggle),
this search results in an approximation that is both monotone and has
derivative values similar to the initial estimates.

>>>>

The quadratic facet model produces a unique sensitivity to input
perturbation, as small changes in input may cause different quadratic
facets to be associated with a breakpoint, and thus different initial
derivative estimates can be produced. This phenomenon is depicted in
Figure 1. Despite this sensitivity, the quadratic facet model is still
preferred because it exactly captures local linear and quadratic
behavior while empirically producing final approximations with less
wiggle ($L^2$ norm of the second derivative) than other methods. A
weighted harmonic mean estimate of first derivatives may be more
accurate when the underlying function changes at a rate greater than a
quadratic, but that method increases the second derivative sensitivity
to small perturbations in data and empirically results in quintic
splines with greater wiggle.

The binary search for a point on the monotone boundary in $(\tau_1$,
$\alpha$, $\beta$, $\gamma)$ space is performed because it results in
monotone quintic spline interpolants with derivative values that are
absolutely nearer to initial estimates than a search that strictly
shrinks derivative values. Given that the initial derivative estimates
have desirable properties (capture low-order phenomena and are low
wiggle), this search results in an approximation that is both monotone
and has derivative values similar to the initial estimates.

### 2021-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Incorporating some most recent comments from Dr. Watson and Thacker

>>>>

Comparisons between this work and lower order spline algorithms are
problematic, especially with sparse data. For exact data from smooth
functions, comparing derivative values, higher order methods always win
as the data density increases. This result is demonstrated in the work of
Lux et al. [2020]. For sparse or synthetic data, what constitutes a
better fit is either subjective (e.g., less wiggle) or dependent on the
specific problem. If an application demands $C^2$ continuity and
monotonicity, the target of {\tt MQSI}, then this software provides such
solutions while no existing published spline algorithms suffice.
